Does stomatal patterning in amphistomatous leaves minimize the CO\(_2\) diffusion path length within leaves?

### 2024-01-22 ms/response-to-reviewers.qmd

Dear authors,

My apologies for the delay in returning this manuscript, as you can understand it can be difficult to find reviewers at this time of year. We have now received two comprehensive reviews of your manuscript. Both reviewers found the science to be interesting and thought provoking. I agree, it was a very interesting read. However, reviewer 1 found it difficult to understand some of the terminology and methods and reviewer 2 has requested clarity on some of the assumptions in your modeling. Both have provided detailed comments for you to address. Could you please revise the manuscript and address their comments before resubmission to AoBP.

Kind regards, William

```{=tex}
\textcolor{blue}{Thank you so much for organizing a review of this manuscript. The reviewer's comments have been extremely useful in improving the manuscript for publication in \textit{AoBP}. We appreciate your help in getting this science in print. We summarize our changes here; details on specific comments can be found below.\\
\\
We have clarified our terminology, adding in definitions where needed and standardizing all uses of specific terms. We've clarified our methodology by being more explicit about which data are real and which are synthetic. We have added all the references and relevant discussion of them as suggested by reviewer 1. We've clarified the wording of our hypotheses following the suggestion of reviewer 1. And, as was addressed by both reviewers 1 and 2, we've clarified the limitations of our model assumptions and extensively addressed them in the introduction and discussion. To accommodate some of these changes, we've made some minor changes elsewhere in the text; however, the largest changes are all reflected in this document. We hope that, with these significant clarifications, changes, and additions of important caveats, our manuscript will be considered for publication in \textit{AoBP}. Thank you for your time and consideration.}
```
## Response to Reviewer #1

@@

There are several descriptions used that are not entirely clear. What exactly is meant by "over dispersal" in the context of stomatal patterning. Does this mean there are too many stomata on the leaf or their spacing is not optimal but the correct number? This is not common terminology when referring to stomatal patterning. The assessment of distribution relative to a uniform random distribution needs further explanation as it is well established that stomata are randomly distributed.

**Response:** \textcolor{blue}{We appreciate the reviewer's time to read and comment on our manuscript. They have brought up important points of clarification that must be addressed before the manuscript is ready to be published.}

\textcolor{blue}{We have clarified and standardized our terminology, adding explanations when needed. It has long been known that stomata are non-randomly distributed, tending to be more uniformly distribution than random (see Croxdale et al. (2000) for a nice review of historical evidence of stomatal patterning). In Croxdale et al. (2000), there is also confusion over terminology in the published literature, so we are not sure that there are field-accepted terms. So, we have adopted our own set of terminology which mirrors that of the field and related fields which discuss the spatial patterns of discrete objects. Specifically, see lines 111-117:}

> "We compared observed stomatal patterning to an ideal pattern (uniform equilateral triangular grid) and a null model (random uniform distribution). The terminology is unfortunately confusing because the word `uniform' is used in different ways. A uniform equilateral triangular grid means that the distance between stomata is uniform; a random uniform distribution means that a stomate has an equal (i.e. uniform) probability of occurring anywhere on the leaf surface. To limit confusion, we refer to the ideal pattern (equilateral triangle grid) as uniform and the null pattern (random uniform) as random. When observed stomatal patterns are more dispersed than expected under random patterning, we refer to this as overdispersed. Note, however, that overdispersed compared to random is still less dispersed than ideal, because the ideal pattern is maximally dispersed."

@@

**Response:** \textcolor{blue}{It now reads: ``pixel of the surface'' on line 156.}

Line 157 - ideal pattern is "hexagonal"?

**Response:** \textcolor{blue}{Thanks for catching this. We’ve now standardized this term to be the same as every other mention of the uniform, equilateral triangular grid. The area of leaf surface closest to a given stomate, when stoamta are arrayed in an equilateral triangular grid, forms a hexagon (see Fig. 2C).}

Line 168 -- "All parameters....exceeded 103" -- please clarify what you mean by this statement.

@@

> "All parameters in the Bayesian linear mixed-effects model converged ($\hat{R} < 1.01$) and effective sample sizes exceeded $10^3$"

## Response to Reviewer #2

Dear authors,

The authors made a theoretical analysis of the optimal distribution of stomata on each leaf surface and then compared it with actual data. I only have a few comments, most of them concerning the assumptions. I think the consequences of their assumptions could be better addressed and explained with further details, especially the concept of "optimal distribution".

### 2024-01-22 ms/ms.Rmd
fixed some typos in revised ms and response to reviewers

    affiliation: NIAB

@@

We compared observed stomatal patterning to an ideal pattern (uniform equilateral triangular grid) and a null model (random uniform distribution). The terminology is unfortunately confusing because the word `uniform' is used in different ways. A uniform equilateral triangular grid means that the distance between stomata is uniform; a random uniform distribution means that a stomate has an equal probability (i.e. uniform) of occurring anyone on the leaf surface. To limit confusion, we refer to the ideal pattern as uniform (equilateral triangle grid) and the null pattern as random (random uniform). When observed stomatal patterns are more dispersed than expected under random patterning, we refer to this as overdispersed. Note, however, that overdispersed compared to random is still less dispersed than ideal, because the ideal pattern is maximally dispersed.

>>>>

    affiliation: NIAB g

@@

We compared observed stomatal patterning to an ideal pattern (uniform equilateral triangular grid) and a null model (random uniform distribution). The terminology is unfortunately confusing because the word `uniform' is used in different ways. A uniform equilateral triangular grid means that the distance between stomata is uniform; a random uniform distribution means that a stomate has an equal (i.e. uniform) probability of occurring anywhere on the leaf surface. To limit confusion, we refer to the ideal pattern (equilateral triangle grid) as uniform and the null pattern (random uniform) as random. When observed stomatal patterns are more dispersed than expected under random patterning, we refer to this as overdispersed. Note, however, that overdispersed compared to random is still less dispersed than ideal, because the ideal pattern is maximally dispersed.

### 2024-01-22 ms/ms.tex
fixed some typos in revised ms and response to reviewers

\author[%
\addrlblbycode{NIAB}%
%
%
%
]{Graham J. Dow}

\addrbycode{NIAB}

@@

\section{Introduction}\label{introduction}

@@

\section{Materials and methods}\label{materials-and-methods}

\subsection{Data Preparation}\label{data-preparation}

@@

\subsection{Single surface analyses}\label{single-surface-analyses}

We compared observed stomatal patterning to an ideal pattern (uniform
equilateral triangular grid) and a null model (random uniform
distribution). The terminology is unfortunately confusing because the
word `uniform' is used in different ways. A uniform equilateral
triangular grid means that the distance between stomata is uniform; a
random uniform distribution means that a stomate has an equal
probability (i.e.~uniform) of occurring anyone on the leaf surface. To
limit confusion, we refer to the ideal pattern as uniform (equilateral
triangle grid) and the null pattern as random (random uniform). When
observed stomatal patterns are more dispersed than expected under random
patterning, we refer to this as overdispersed. Note, however, that
overdispersed compared to random is still less dispersed than ideal,
because the ideal pattern is maximally dispersed.

@@

\noindent where \(d_i\) is the distance between \(\text{stomate}_i\) and
its nearest neighbor. We calculated \(\mathrm{NNI}\) using the \emph{R}
package \textbf{spatialEco} version 2.0.2 \citep{evans_spatialeco_2023}.
The observed stomatal distribution is overdispersed relative to a random
distribution if the observed \(NNI\) is greater than 95\% of the
synthetic \(\mathrm{NNI}\) values (one-tailed test). We adjusted
\(P\)-values to account for multiple comparisons using the
Benjamini-Hochberg \citep{benjamini_controlling_1995} false discovery
rate procedure implemented in the \emph{R} package \textbf{multtest}
version 2.56.0 \citep{wong_multiple_2005}.

@@

\subsection{Paired Abaxial and Adaxial Surface
Analysis}\label{paired-abaxial-and-adaxial-surface-analysis}

@@

\subsection{Modeling Photosynthesis}\label{modeling-photosynthesis}

@@

\section{Results}\label{results}

@@

\subsection{Stomatal patterning is non-random, but far from
uniform}\label{stomatal-patterning-is-non-random-but-far-from-uniform}

@@

\subsection{No evidence for coordinated stomatal position between
surfaces}\label{no-evidence-for-coordinated-stomatal-position-between-surfaces}

@@

\subsection{Larger stomata supply larger mesophyll
volumes}\label{larger-stomata-supply-larger-mesophyll-volumes}

@@

\subsection{Little benefit of coordinated stomatal
arrangement}\label{little-benefit-of-coordinated-stomatal-arrangement}

@@

\section{Discussion}\label{discussion}

>>>>

\author[%
\addrlblbycode{NIAB g}%
%
%
%
]{Graham J. Dow}

\addrbycode{NIAB g}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

\hypertarget{materials-and-methods}{%
\section{Materials and methods}\label{materials-and-methods}}

\hypertarget{data-preparation}{%
\subsection{Data Preparation}\label{data-preparation}}

@@

\hypertarget{single-surface-analyses}{%
\subsection{Single surface analyses}\label{single-surface-analyses}}

We compared observed stomatal patterning to an ideal pattern (uniform
equilateral triangular grid) and a null model (random uniform
distribution). The terminology is unfortunately confusing because the
word `uniform' is used in different ways. A uniform equilateral
triangular grid means that the distance between stomata is uniform; a
random uniform distribution means that a stomate has an equal
(i.e.~uniform) probability of occurring anywhere on the leaf surface. To
limit confusion, we refer to the ideal pattern (equilateral triangle
grid) as uniform and the null pattern (random uniform) as random. When
observed stomatal patterns are more dispersed than expected under random
patterning, we refer to this as overdispersed. Note, however, that
overdispersed compared to random is still less dispersed than ideal,
because the ideal pattern is maximally dispersed.

@@

\noindent where \(d_i\) is the distance between \(\text{stomate}_i\) and
its nearest neighbor. We calculated \(\mathrm{NNI}\) using the \emph{R}
package \textbf{spatialEco} version 2.0.2 \citep{evans_spatialeco_2023}.
The observed stomatal distribution is overdispersed relative to a random
distribution if the observed \(NNI\) is greater than 95\% of the
synthetic \(\mathrm{NNI}\) values (one-tailed test). We adjusted
\(P\)-values to account for multiple comparisons using the
Benjamini-Hochberg \citep{benjamini_controlling_1995} false discovery
rate procedure implemented in the \emph{R} package \textbf{multtest}
version 2.58.0 \citep{wong_multiple_2005}.

@@

\hypertarget{paired-abaxial-and-adaxial-surface-analysis}{%
\subsection{Paired Abaxial and Adaxial Surface
Analysis}\label{paired-abaxial-and-adaxial-surface-analysis}}

@@

\hypertarget{modeling-photosynthesis}{%
\subsection{Modeling Photosynthesis}\label{modeling-photosynthesis}}

@@

\hypertarget{results}{%
\section{Results}\label{results}}

@@

\hypertarget{stomatal-patterning-is-non-random-but-far-from-uniform}{%
\subsection{Stomatal patterning is non-random, but far from
uniform}\label{stomatal-patterning-is-non-random-but-far-from-uniform}}

@@

\hypertarget{no-evidence-for-coordinated-stomatal-position-between-surfaces}{%
\subsection{No evidence for coordinated stomatal position between
surfaces}\label{no-evidence-for-coordinated-stomatal-position-between-surfaces}}

@@

\hypertarget{larger-stomata-supply-larger-mesophyll-volumes}{%
\subsection{Larger stomata supply larger mesophyll
volumes}\label{larger-stomata-supply-larger-mesophyll-volumes}}

@@

\hypertarget{little-benefit-of-coordinated-stomatal-arrangement}{%
\subsection{Little benefit of coordinated stomatal
arrangement}\label{little-benefit-of-coordinated-stomatal-arrangement}}

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}
